---
title: "Food insecurity Intervention"
author: "Angel Gallardo"
date: "2025-11-08"
bibliography: library.bib
format: pdf
---

# Abstract

Food insecurity is a global issue, in the United States alone, 13.5% of households or 18 million households were food insecure in 2023. With high estimates of food waste, this research looks to find intervention through collaboration with restaurants and government programs. The goal is to build a model that accurately predicts high food insecurity at the county level. Furthermore, we look to demonstrate the potential impact that local restaurants and government programs can create with efficient intervention. This way solving food insecurity in the United States can also improve food waste from restaurants in the community. Policy making can improve at the county level to help mitigate food insecurity and food loss with food assistance programs. By adding food assistance program participation data into the model we can understand how Americans are securing their food. Thus this research looks to detect whether calorie intake can predict high food insecurity. Finally, an intervention involving the recovery of a percentage of restaurant food loss through SNAP policy is simulated to estimate change in food insecurity.

# Introduction

In 2023, the USDA, FDA, and EPA [@fda2023nationalstrategy]announced a national goal to reduce food loss and waste in the U.S by 50% by 2030. Food insecurity [@feedingamerica_foodinsecurity] is described as people not having access to food needed to live their fullest lives. On the other side, Food waste [@ReFED2025] is defined as the edible amount of food, post harvest, that is available for human consumption but is not yet consumed for any reason. Additionally, over one-third of all available food goes uneaten through loss or waste.

According to Feeding America [@feedingamerica_foodinsecurity], the causes of food insecurity include income-related factors (low-wage jobs, unreliable work, or financial emergencies); high cost of living (essentials are increasingly difficult to afford); community factors (limited access to transportation); health-related issues (reduced budgets for food and other essentials); and systematic barriers (unfair systems in policies and institutions).

By using readily available secondary data such as income levels, community factors, health issues, and combining it with government assistance program participation we can:

Create a model that isolate counties that are most need of intervention, develop a model that can predict high food insecurity using key demographics as well as government program participation rates. Through this model, we will understand the main indicators of food insecurity in the United states. Finally, a potential intervention policy is simulated to analyze the effects it may have on food insecurity.

# Research Question

The focus of this research is at the county level, where a model can reveal high priority counties for food security intervention. Furthermore, the findings could create a collaborative effort between local food retailers and non-profits through incentive-based programs aimed to effectively and efficiently reduce food insecurity. By developing a model to accurately predict high priority counties, food pantries and government programs can rapidly focus on collaborating with local retailers to mitigate hunger. In addition, local retailers may also benefit by reduction of food loss while possible increasing revenue. In essence, this model could be utilize for non-profits to collaborate with local retailers, or the government can use this for policy making to increase food program authorized stores where needed. Currently, each state sets their own requirements for choosing how many and which restaurants they allow to participate in the SNAP program.

The significance of this phenomenon is twofold. First, more than 95 percent of food waste in the United States ends up in landfills, where it breaks down anaerobically and releases methane, carbon dioxide, and other greenhouse gases into the atmosphere [@mu2017environmental]. If food waste is not addressed, it will continue to contribute significantly to climate change. Second, the USDA reported that in 2023, 13.5 percent of U.S. households experienced food insecurity at some point during the year. Food insecurity is a global crisis, and countries around the world are actively seeking solutions to eliminate it.

Addressing food waste and insecurity has numerous real-world applications. For businesses, especially in agriculture, food production, and retail-reducing food waste can lead to cost savings, more efficient supply chains, and improved sustainability practices. Food that would otherwise go to waste can be redistributed to food banks and local communities, helping to reduce hunger in America and improve public health outcomes. Local non-profits and governments can use this model to develop targeted programs and initiatives that connect surplus food with those in need. Taxpayers also benefit when less food waste means better use of public resources and reduced environmental damage. Furthermore, reducing food waste directly contributes to the fight against climate change by lowering greenhouse gas emissions and easing the strain on natural resources. In the long term, tackling this issue supports environmental sustainability, economic efficiency, and social equity.

**Research question:** Can a machine learning model accurately identify food insecurity by US counties? Which indicators are significant in predicting food insecurity? Finally, can a food intervention policy to recover restaurant food loss reduce high food insecurity counties?

I hypothesize that food recovery from restaurants may help reduce food insecurity in mainly rural areas and where there is low access to transportation or stores.

# Literature Review

Measuring Food insecurity and Hunger in the United States [@carlson1999measuring] developed measuring scale to quantify food insecurity in the US. The measurement scale included household surveys where the questions were categorically measured to reflect the severity of the responses. This benchmark classifies food security to be defined as assured access at all times to enough food for an active healthy life. The other form of food insecurity, malnutrition, is not the case in the US, as there are no clinical symptoms of deprivation. As such, questions from skipping meals or children being hungry estimated a higher severity of food insecurity. The results show that severe hunger was more prevalent in households with children under eighteen years old. Additionally, the scale identified that food insecurity was more common among minority households and low income to poverty household ratios. This scale serves as a baseline to understand the relationship between food insecurity and key demographics. However, a predictive model can utilize these predictors, as well as economic access, and calorie intake data to estimate the probability that a household becomes food insecured.

One research paper [@lentz2019data] pointed out that the Integrated Food Security Phase Classification (IPC), a global framework that classifies food insecurity, does not make use of the readily available secondary data, in a replicable and transparent manner. Furthermore, there is no current food security early warning and monitoring system that incorporates readily available data into a predictive model. As a result, these methods limit timely, targeted humanitarian response. The research builds linear and log-linear models using three readily available secondary data measures, weather and price fluctuations, geographic variations, demographics, and intra-annual seasonal trends to predict food security at the sub national level. Their findings show that their model correctly predicts 83% to 99% of the most food insecure clusters, while the IPC’s approach only predicts between 0% and 10%. Although the model is an improvement from the IPC's approach, the model does not incorporate solutions to reduce food insecurity. Meanwhile, the U.S, unlike other countries, weather changes or geographic variations may not be useful as food insecurity is mainly driven by food program access, unemployment, and household characteristics. In addition, by accurately predicting high food insecurity, the US government can assess whether counties are steering towards high food insecurity to perform early intervention.

Leveraging Machine Learning for Food Waste Reduction: An Analysis of Predictive Models [@yang2024foodwaste] predicted food waste using data on household consumption patterns, retail demand, and food service estimates; along with confidence in estimates to gauge reliability for each country. It focuses on machine learning algorithms like Random Forest to predict food waste patterns for support policymakers in developing policies that facilitate interventions to reduce food loss. The report concluded that household-level food waste estimates are the most influential factors driving food waste predictions at the per-capita level. However, household data may be hard to collect due to the reliability of the self-reported data. In addition, collecting reliable household data remains challenging due to the limitations of self-reported information and the lack of incentives for households to participate in surveys. As a result, using household food excess as a solution may be less practical or effective to redistribute excess food.

Another study [@irani2018managing] examines how food waste and loss in supply chains impact food security, using causal modeling and simulation to identify key drivers from qualitative data. The findings show that food waste is not solely the result of kitchen inefficiencies or over-ordering, but arises from a combination of factors, including consumer behavior, storage logistics, and supplier practices. However, one study was noted as identifying plate waste to compomise the highest amount of food waste, followed by storage, and and preperation loss. The research employs a Fuzzy Cognitive Map (FCM) to model the interrelated factors driving food waste and loss. The study concludes that no single factor—such as inventory management—can reduce food waste effectively without complementary actions, such as menu planning and staff training. The study’s reliance on qualitative data and prototype simulations limits its generalizability, and restaurants may need context-specific data to accurately quantify the impact of interventions. It is estimated that 17% of food is wasted at the retail level, where redistribution can have a bigger impact than at the production or consumption stages of the supply chain. Although food waste at the retail level has been studied less than at other supply chain stages, retail stores tend to generate more food waste than food loss, as the discarded food is generally still consumable.

 

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(usmap)
library(ggplot2)
library(scales)
library(Metrics) 
library(caret)
library(randomForest)
library(stringr)
library(jsonlite)
library(httr)
library(purrr)
library(skimr)
library(GGally)
library(ggforce)
library(corrplot)
```

## Data

## Food insecurity

The food environment atlas [@USDA_Food_Environment_Atlas] provides secondary data by state and county. The data set includes the food insecurity by state only, demographics, and food-program participation rates. For our model we will include the makeup of the counties race and age, the median income, obesity rate, and food programs participation rates.

The re-evaluation of the Thrifty food Plan [@USDA_ThriftyFoodPlan2021] in 2021 to estimate the cost of a practical, nutritious, and budget-conscious diet. In 2021 the food plan referenced a family of four, where the cost of the market basket was calculated at \$835.57 per month. The referenced family consisted of two adults, male and female, between the ages of 20-50 years old, calculated a daily intake of calories of about 3,000 and 2,200. The other family members included two children, one with the age rage 6-8, and a daily calorie intake of 1,800, and the other in the 9-11 age category with a daily intake of 2,200. This means that on average one dollar of SNAP benefit purchases 330 calories. This information will be utilized to create additional variable- calories per SNAP recipient- where we can visualize whether the adequate allotment is given to participants.

## Restaurant excess estimates

Food loss and waste data is scarce due to the uneven information across all stages of food chains, and aggregated data that is difficult to replicate. In fact, a UNEP study [@SolohaDace2025] noted that the food waste per capita in the EU cannot be fully explained because the methodology used to quantify food waste in each member state is not known, making it difficult to conclude whether the differences in data are justified or are due to differences in scope. One review paper, [@xue2017missing] identified that data is heavily biased towards industrialized and high-income countries. The paper also points out that fewer studies focus on retail, households, and distribution levels of the supply chain.

US EPA [@EPA_ExcessFoodOpportunitiesMap] provides a map of restaurants at the county level with excess food estimates. We will use this data to find the aggregate food loss in tons from restaurants at the county level. This data will be utilized to estimate the amount of available calories that can be recoverable at the local level. Finally, a percentage of the recoverable amount will be estimated and simulated to calculate the impact it has to reduce food insecurity.

```{r}
insecurity <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/StateAndCountyData.csv")

fi <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/2023%20County%20Health%20Rankings%20Data%20-%20v2.csv")

Restaurant <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/RestaurantsAndFoodServices.csv")

```

```{r}
glimpse(fi)
```

```{r}
glimpse(insecurity_wider)
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
insecurity_wider <- insecurity %>%
  pivot_wider(.,names_from = Variable_Code, values_from = Value)

#glimpse(insecurity_wider)
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
insecurity_wider <- insecurity_wider %>%
  
  mutate(across(where(is.numeric), ~replace(., . %in% c(-8888, -9999), NA))) %>%
  
  mutate(
    PERPOV17_21 = as.factor(replace(PERPOV17_21, PERPOV17_21 %in% c(-8888, -9999), NA)),
    METRO23 = as.factor(replace(METRO23, METRO23 %in% c(-8888, -9999), NA)),
    POPLOSS15 = as.factor(replace(POPLOSS15, POPLOSS15 %in% c(-8888, -9999), NA))
  )
```

```{r}
state_lookup <- data.frame(
  State = state.name,
  StateAbbrev = state.abb,
  stringsAsFactors = FALSE
)

fi <- fi %>%
  left_join(state_lookup, by = c("State" = "State")) %>%
  mutate(State = StateAbbrev) %>%       # replace with abbreviations
  select(-StateAbbrev)

fi <- fi %>% 
  select(State, County, Food_insecure = X..Food.Insecure.1  ,Population, Household_burden = X..Households.with.Severe.Cost.Burden.1, Rural_percent =X..Rural  )
insecurity_wider <- insecurity_wider %>%
  left_join(fi, by = c("State", "County"))

glimpse(insecurity_wider)
```

additional SNAP variables

```{r}
insecurity_wider <- insecurity_wider %>% 
  mutate( total_SNAP = PC_SNAPBEN22*Population,
          SNAP_participants = Population* (PCT_SNAP22/100),
          SNAP_per_participant= (total_SNAP/SNAP_participants)
          )
#glimpse(insecurity_wider)   
```

add calories per month

```{r}

# add calories per month

insecurity_wider <- insecurity_wider%>%
  mutate(calories_SNAP_part = SNAP_per_participant * 330)
```

```{r}

```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
Restaurant <- Restaurant %>%
  mutate(Excess_food = 
           (`Excess.Food.Estimate..Low..tons.per.year.` +
              `Excess.Food.Estimate..High..tons.per.year.`) / 2) %>%
  group_by(State, County) %>% 
  summarise(
    Total_Excess_Food = sum(Excess_food, na.rm = TRUE),
    n_restaurants = n()  

  ) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

glimpse(Restaurant)
```

```{r}
#combine data
combined<- insecurity_wider %>%
  left_join(Restaurant, by = c("State", "County"))




#Estimated calories wasted per capita
combined <- combined %>%
  mutate(kcal_wasted_per_capita = (Total_Excess_Food*2000*1400)/Population,
         restaurants_per_capita = n_restaurants/Population)

glimpse(combined)

```

```{r}
# use relevane variables
combined <- combined %>%
  select( State, County,Food_insecure ,PCT_NHWHITE20, PCT_NHBLACK20, PCH_WICWOMEN_16_21, PCT_HISP20, PCT_NHTMR20, MEDHHINC21, PCH_SNAP_17_22, PCH_LACCESS_LOWI_15_19, PCH_LACCESS_HHNV_15_19, PCT_OBESE_ADULTS22, PCH_SFSP_17_21, PCH_WIC_17_21, REDEMP_SNAPS23,PC_SNAPBEN22,REDEMP_WICS22,SNAPSPTH23, PCT_LACCESS_HHNV19, PCH_SNAPSPTH_17_23, PCT_LACCESS_SNAP19, POVRATE21, METRO23,total_SNAP, SNAP_participants , SNAP_per_participant, calories_SNAP_part ,PCT_65OLDER20, PCT_18YOUNGER20,Population, Household_burden, Rural_percent,Total_Excess_Food , n_restaurants, kcal_wasted_per_capita, restaurants_per_capita
 )

glimpse(combined)
```

# Data Exploration

The data summary shows that most of the variables data is complete, except for the WIC redemption variable at 55% complete. PCH_LACCESS_LOWI_15_19, defined as low income and low access to stores between 2015-2019 has a large standard deviation which could be due to population migration. The mean food insecurity rate per county is at 12.5%, with a standard deviation of 3%. The highest percent of food insecurity is at 29%. As expected, the white population per county is right-skewed, meaning that most counties has high white population. The mean poverty rate in the US is at 14.5%, where some counties experience poverty rates as high as 44%.

```{r}
options(scipen = 999)   

skim(combined)
```

The histogram plots demonstrate that most of the variables are either left or right skewed. For instance, the the histogram shows most counties have a high percentage of white residents, but there are extreme cases where counties have a low white population. However, food insecurity, poverty, population of residents under the age of 18, and over 65 are normally distributed. Because the dependent variable, food insecurity, is approximately normally distributed, it is more compatible with linear regression assumptions and may improve model accuracy. The skewed independent variables show that a transformation or a model robust to non-normal predictors may be more appropriate. In addition, the scatter plots confirm that most predictors exhibit heteroscedasticity against food insecurity, meaning that the variance is not consistent across the range of predictors. Overall, the the skewed predictors, as well as the presence of outliers indicate that variable transformation or non-linear models are more adequate to predict food insecurity.

```{r}
numeric_vars<- names(combined)[sapply(combined, is.numeric)]
combined_long <- combined %>%
  pivot_longer(cols = all_of(numeric_vars),
               names_to = "variable",
               values_to = "value")

ggplot(combined_long, aes(x = value)) +
  geom_histogram(bins = 20) +
  facet_wrap_paginate(~ variable, scales = "free", ncol = 5, nrow = 5, page = 1)
```

```{r}

numeric_vars <- setdiff(names(combined)[sapply(combined, is.numeric)], "Food_insecure")

combined_long <- combined %>%
  pivot_longer(cols = all_of(numeric_vars),
               names_to = "variable",
               values_to = "value")

ggplot(combined_long, aes(x = value, y = Food_insecure)) +
  geom_point(alpha = 0.5, color = "blue") +
  facet_wrap_paginate(~ variable, scales = "free_x", ncol = 5, nrow = 4, page = 1) +
  theme_minimal() +
  labs(x = "Variable Value", y = "Food Insecure",
       title = "Predictors vs Food Insecure")

```

The top predictors highly correlated with food insecurity are poverty, median income,

```{r}
corr_with_foodinsec <- combined %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ cor(.x, combined$Food_insecure,
                                       use = "pairwise.complete.obs"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))

corr_with_foodinsec
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
top_vars <- corr_with_foodinsec %>%
  slice_max(order_by = abs(Correlation), n = 15) %>%
  pull(Variable)
top_vars <- setdiff(top_vars, "Food_insecure")

top_data <- combined %>%
  select(all_of(top_vars))


ggpairs(top_data,
        upper = list(continuous = wrap("cor", size = 2)),
        lower = list(continuous = wrap("points", alpha = 0.5, size = 1.5)),
        diag = list(continuous = wrap("barDiag", fill = "lightblue"))
) + 
theme(strip.text = element_text(face = "bold",size = 7,  angle = 50))
```

# Data Preparation

Remove outliers, Food insecurity as a binary variable to predict high insecurity.

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}


```

```{r}
combined <- combined %>%
  mutate(High_FoodInsecure = ifelse(Food_insecure >= 12, 1, 0))

```

```{r}
Res_price_per_cal <- .02
discount_rate <-1.10
calories_per_ton <- 1652000
additional_snap_spend <- 100

combined < combined %>%
  mutate(calories_wasted = Total_Excess_Food * calories_per_ton )%>%
  mutate(calories_cost = calories_wasted * Res_price_per_cal) %>%
  mutate(Recovered_calories =  calories_cost/ (additional_snap_spend* discount_rate))

```

```{r}
mode_impute <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

combined <- combined %>%
  group_by(State) %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()


combined <- combined %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Winsorize to remove outliers
winsorize <- function(x, probs = c(0.01, 0.99)) {
  q <- quantile(x, probs = probs, na.rm = TRUE)
  pmin(pmax(x, q[1]), q[2])
}

combined <- combined %>%
  mutate(across(where(is.numeric), winsorize))


combined <- combined %>%
  mutate(
    METRO23 = as.numeric(as.character(METRO23)),  
    METRO23 = as.numeric(mode_impute(METRO23))     
  )

na_counts <- colSums(is.na(combined))
na_columns <- names(na_counts[na_counts > 0])
na_columns

head(combined)
```

```{r}

check <- combined %>%
  select(where(~ any(is.na(.))))

summary(check)


```

```{r}

cat(names(combined), sep = "\n")

# Only numeric variables as a vertical list
numeric_vars <- names(combined)[sapply(combined, is.numeric)]
cat(numeric_vars, sep = "\n")

# Only factor/character variables as a vertical list
factor_vars <- names(combined)[sapply(combined, function(x) is.factor(x) | is.character(x))]
cat(factor_vars, sep = "\n")
```

```{r}
## rename columns for clarity
combined <- combined %>%
  rename(
    Foodinsecurity = Food_insecure,
    WhitePct = PCT_NHWHITE20,
    BlackPct = PCT_NHBLACK20,
    HispPct = PCT_HISP20,
    TwoOrMorePct = PCT_NHTMR20,
    Medianincome = MEDHHINC21,
    low_income_low_store_access = PCH_LACCESS_LOWI_15_19,
    ObesityPct = PCT_OBESE_ADULTS22,
    WICChange = PCH_WIC_17_21,
    SNAPparticipantschange = PCH_SNAP_17_22,
    SNAPstoreschange = PCH_SNAPSPTH_17_23,
    SummerFoodProgChange = PCH_SFSP_17_21,
    MetroStatus = METRO23,
    adult_obese_rate = PCT_OBESE_ADULTS22,
    WomenWICpart = PCH_WICWOMEN_16_21,
    Poverty_rate = POVRATE21,
    WIC_redemptions = REDEMP_WICS22,
    SNAP_redemptions = REDEMP_SNAPS23,
    SNAP_per_capita = PC_SNAPBEN22,
    SNAP_auth_stores = SNAPSPTH23,
    SNAP_low_access_pct = PCT_LACCESS_SNAP19,
    Pct_65_older = PCT_65OLDER20,
    PCT_18_younger = PCT_18YOUNGER20,
    No_car_low_store = PCH_LACCESS_LOWI_15_19 
  )

model_data <- combined %>%
  select(-State, -County,-Foodinsecurity )
 
glimpse(model_data)

```

```{r}
corr_with_foodinsec2 <- model_data%>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ cor(.x, combined$High_FoodInsecure,
                                       use = "pairwise.complete.obs"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))

corr_with_foodinsec2
  
```

```{r}
# 1. Select numeric variables
numeric_vars2 <- model_data %>% select(where(is.numeric))

# 2. Compute correlations with Foodinsecurity
cor_with_food <- sapply(numeric_vars2, function(x) cor(x, numeric_vars2$Foodinsecurity, use = "complete.obs"))

# 3. Convert to data frame
cor_df <- data.frame(
  Variable = names(cor_with_food),
  Correlation = round(cor_with_food, 2)
) %>%
  filter(Variable != "Food_insecure") %>%   # remove self-correlation
  arrange(desc(abs(Correlation)))

# 4. View table
cor_df

# 5. Plot correlations
ggplot(cor_df, aes(x = reorder(Variable, Correlation), y = Correlation)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Correlation with Foodinsecurity",
       x = "Variable",
       y = "Correlation") +
  theme_minimal()
```

```{r}
numeric_vars2 <- names(model_data)[sapply(model_data, is.numeric)]
scatter_plots2 <- lapply(numeric_vars2, function(col_name){
  ggplot(model_data, aes_string(x = col_name, y = "Foodinsecurity")) +
    geom_point(alpha = 0.6, color = "blue") +
    labs(title = paste("Food insecurity vs", col_name),
         x = col_name,
         y = "Foodinsecure") +
    theme_minimal()
})
scatter_plots2
```

```{r}
data_plot <- lapply(names(model_data), function(col_name){
  ggplot(model_data, aes_string(x =col_name)) +
    geom_histogram(fill = "red", bins = 20, na.rm = TRUE) +
    labs(title = col_name, x = col_name, y = "count") +
    theme_minimal()    
})

data_plot


```

# Models

## Methodology

This study models county level food insecurity based on demographics, economic factors, health, and government program participation. Prior research utilizes demographic statistics such as race and genders to understand the ethnic groups that may contribute the unequal access to food. This is especially true in the United States as the racial demographic of the United states is very diversified. undefined

Government assistance indicators are also used as predictors. Changes in participation in SNAP, WIC, the Summer Food Program, and other support programs provide insight into how resource availability and policy implementation may influence local food access. These variables serve as proxies for demand for assistance and the adequacy of food support infrastructure.

Socioeconomic conditions are represented by median household income, which is a strong determinant of food affordability and financial stability. Finally, county-level obesity rates are included because diet-related health outcomes often correlate with limited access to healthy and affordable foods. Together, these variables form a comprehensive predictive model for understanding the drivers of food insecurity across U.S. counties.

## linear model

```{r}

set.seed(1234)


sample<- sample(nrow(model_data), nrow(model_data) *.60,
                replace = FALSE)

```

```{r}
foodinsec_train <- model_data[sample,]
foodinsec_train

foodinsec_test <-  model_data[-sample,]
foodinsec_test
```

```{r}
# modeling all predictors
model1 <- glm(
  High_FoodInsecure ~ Poverty_rate + Medianincome + calories_SNAP_part+ WhitePct + TwoOrMorePct + BlackPct + HispPct + SNAP_low_access_pct + Rural_percent +  Pct_65_older + PCT_18_younger +restaurants_per_capita
 ,
  data = foodinsec_train,
  family = binomial
)
summary(model1)
```

```{r}



predictions <- predict(model1, newdata = foodinsec_test)

rmse_test <- rmse(foodinsec_test$Foodinsecurity, predictions)

rmse_test


```

```{r}
# model 2
set.seed(12345)
lm_model2 <- lm(
  Foodinsecurity ~ BlackPct + AsianPct + WomenWICpart + NativePct + HispPct + TwoOrMorePct +
                   Medianincome + low_income_low_store_access + adult_obese_rate + 
                   SummerFoodProgChange + WICChange + SNAP_redemptions + SNAP_per_capita +
                   WIC_redemptions + SNAP_auth_stores + SNAPstoreschange +
                    calories_per_person,
  data = foodinsec_train
)
summary(lm_model2)
```

```{r}
predictions2 <- predict(lm_model2, newdata = foodinsec_test)

rmse_test2 <- rmse(foodinsec_test$Foodinsecurity, predictions)

rmse_test2
```

```{r}
AIC(model1, lm_model2)
BIC(model1, lm_model2)
```

## Random Forest

```{r}
set.seed(111)
rf_model <- randomForest(
  Foodinsecurity ~ WhitePct + BlackPct + NativePct + 
    PacificIslanderPct +  Medianincome + NocarlowaccessChange + 
    ObesityPct + WICChange + SNAPparticipantschange + 
    SNAPstoreschang + SummerFoodProgChange + MetroStatus + 
    WomenWICpart,
  data = foodinsec_train,
  ntree = 500,
  importance = TRUE
)

print(rf_model)
importance(rf_model)
varImpPlot(rf_model)
```

```{r}
set.seed(1235)
control <- trainControl(method = "cv", number = 5)
rf_cv <- train(
  Foodinsecurity ~ WhitePct + BlackPct + NativePct + PacificIslanderPct +
    Medianincome + NocarlowaccessChange + ObesityPct + WICChange +
    SNAPparticipantschange + SNAPstoreschang + SummerFoodProgChange +
    MetroStatus + WomenWICpart,
  data = model_data,
  method = "rf",
  trControl = control,
  tuneLength = 3
)
print(rf_cv)
```

# Conclusion

The first model developed to predict food insecurity by county is a multiple linear regression which included key demographic variables, as well as economic, access to stores, and participation in food programs. This model scored an adjusted R\^2 of 71%. Most variables were statistically significant, except for the Asian population in the county and the adult obesity rate. As expected, median income, various types of food programs participation changes, and overall obesity rates were statistically significant. This indicates that some of the variables such as adult obesity rate was redundant and some multicollinearity.

The second model removed the weaker predictors, due to the insignificance to the model. However, the adjusted R\^2 remained about the same at \~71%. This concluded that the excluded variables were in fact not valuable to the model.

Lastly, the random forest model used the same predictors as the previous model to capture any nonlinear relationships among the variables. This modeled outperformed with an R\^2 of 99.97%, an almost perfect accuracy rate. WIC participation from women was the most influential variable, which is as expected because this variable had a high correlation with food insecurity.

```{r}
model_summary <- data.frame(
  Model = c("Linear Regression (Full)", 
            "Linear Regression (Reduced)", 
            "Random Forest"),
  RMSE = c(1.295, 1.29, 0.0385),
  R2 = c(0.7091, 0.7091, 0.99976),
  MAE = c(NA, NA, 0.0105)  
)

model_summary

```

```{r}
model_long <- model_summary %>%
  pivot_longer(cols = c(RMSE, R2), names_to = "Metric", values_to = "Value")


ggplot(model_long, aes(x = Model, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("R2" = "#1b9e77", "RMSE" = "#d95f02")) +
  labs(title = "Model Performance Comparison",
       x = "Model",
       y = "Metric Value") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
ggplot(model_long, aes(x = reorder(Model, -Value), y = Value, fill = Metric)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(title = "Model Comparison: Accuracy and Error",
       x = "Model", y = "Value") +
  theme_minimal(base_size = 13)
```
