---
title: "Food insecurity Intervention"
author: "Angel Gallardo"
date: "2025-11-08"
bibliography: library.bib
format: pdf
---

# Abstract

Food insecurity is a global issue, in the United States. 13.5% of households or 18 million households were food insecure in 2023. With high estimates of food waste, this research looks to find intervention through collaboration with restaurants and government programs. The goal is to build a model that accurately predicts high food insecurity at the county level. Furthermore, we look to demonstrate the potential impact that local restaurants and government programs can make with early intervention. Solving food insecurity in the United States can also improve food waste from restaurants in the community. Policy making can improve at the county level to help mitigate food insecurity and food loss with food assistance programs. Food programs participation changes provides important data on how Americans are securing their food. Thus this model relies heavily on various government assistance programs to understand food insecurity.

# Introduction

In 2023, the USDA, FDA, and EPA [@fda2023nationalstrategy]announced a national goal to reduce food loss and waste in the U.S by 50% by 2030. Food insecurity [@feedingamerica_foodinsecurity] is described as people not having access to food needed to live their fullest lives. Food waste [@ReFED2025] is defined as the edible amount of food, post harvest, that is available for human consumption but is not yet consumed for any reason. According to USDA, "over one-third of all available food goes uneaten through loss or waste.

According to Feeding America [@feedingamerica_foodinsecurity], the causes of food insecurity include income-related factors (low-wage jobs, unreliable work, or financial emergencies); high cost of living (essentials are increasingly difficult to afford); community factors (limited access to transportation); health-related issues (reduced budgets for food and other essentials); and systematic barriers (unfair systems in policies and institutions).

By using readily available secondary data such as income levels, community factors, health issues, and combining it with government assistance program participation we can:

Create a priority score model that isolate counties that are most need of intervention

Develop a model that can predict high food insecurity using key demographics as well as government program participation rates.

Determine what the main causes of food insecurity in the United states

Find counties with high levels of food insecurity and restaurant food loss for immediate intervention

# Research Question

The focus of this research is at the county level, where a model can reveal high priority counties for food security intervention. Furthermore, the findings could create a collaborative effort between local food retailers and non-profits through incentive-based programs aimed to effectively and efficiently reduce food insecurity. By developing a model to accurately predict high priority counties, food pantries and government programs can rapidly focus on collaborating with local retailers to mitigate hunger. In addition, local retailers will also benefit by reduction of food loss while possible increasing revenue. In essence, this model could be utilize for non-profits to collaborate with local retailers, or the government can use this for policy making to increase food program authorized stores where needed. Currently, each state sets their own requirements for choosing how many and which restaurants they allow to participate in the SNAP program.

The significance of this phenomenon is twofold. First, more than 95 percent of food waste in the United States ends up in landfills, where it breaks down anaerobically and releases methane, carbon dioxide, and other greenhouse gases into the atmosphere [@mu2017environmental]. If food waste is not addressed, it will continue to contribute significantly to climate change. Second, the USDA reported that in 2023, 13.5 percent of U.S. households experienced food insecurity at some point during the year. Food insecurity is a global crisis, and countries around the world are actively seeking solutions to eliminate it.

Addressing food waste and insecurity has numerous real-world applications. For businesses, especially in agriculture, food production, and retail-reducing food waste can lead to cost savings, more efficient supply chains, and improved sustainability practices. Food that would otherwise go to waste can be redistributed to food banks and local communities, helping to reduce hunger in America and improve public health outcomes. Local non-profits and governments can use this model to develop targeted programs and initiatives that connect surplus food with those in need. Taxpayers also benefit when less food waste means better use of public resources and reduced environmental damage. Furthermore, reducing food waste directly contributes to the fight against climate change by lowering greenhouse gas emissions and easing the strain on natural resources. In the long term, tackling this issue supports environmental sustainability, economic efficiency, and social equity.

**Research question:** Can a predictive priority-score/model model identify US counties where redistributing restaurant excess food will most effectively reduce county-level food insecurity?

# Literature Review

There has not been much research on how to solve food insecurity through food waste estimates. For instance, one research paper [@lentz2019data] pointed out that the Integrated Food Security Phase Classification (IPC), a global framework that classifies food insecurity, does not make use of the readily available secondary data, in a replicable and transparent manner. Furthermore, there is no current food security early warning and monitoring system that incorporates readily available data into a predictive model. As a result, these methods limit timely, targeted humanitarian response. The research builds linear and log-linear models using three readily available secondary data measures, weather and price fluctuations, geographic variations, and intra-annual seasonal trends to predict food security at the sub national level. Their findings show that their model correctly predicts 83% to 99% of the most food insecure clusters, while the IPC’s approach only predicts between 0% and 10%. Although the model is an improvement from the IPC's approach, the model does not incorporate solutions to reduce food insecurity. Meanwhile, the U.S, unlike other countries, has scorches of useful readily available secondary data to predict priority counties. In addition, by accurately predicting priority zones, the US government can assess whether counties are steering towards high food insecurity to perform early intervention.

Leveraging Machine Learning for Food Waste Reduction: An Analysis of Predictive Models [@yang2024foodwaste] predicts food waste using data on household consumption patterns, retail demand, and food service estimates; along with confidence in estimates to gauge reliability for each country. It focuses on machine learning algorithms like Random Forest to predict food waste patterns for informed strategies. The report concluded that household-level food waste estimates are the most influential factors driving food waste predictions at the per-capita level. However, household data may be hard to collect due to the reliability of the self-reported data. In addition, collecting reliable household data remains challenging due to the limitations of self-reported information and the lack of incentives for households to participate in surveys. As a result, using household food excess as a solution may be less practical or effective to redistribute excess food.

Another study [@irani2018managing] examines how food waste and loss in supply chains impact food security, using causal modeling and simulation to identify key drivers. The findings show that food waste is not solely the result of kitchen inefficiencies or over-ordering, but arises from a combination of factors, including consumer behavior, storage logistics, and supplier practices. The research employs a Fuzzy Cognitive Map (FCM) to model the interrelated factors driving food waste and loss. However, the study concludes that no single factor—such as inventory management—can reduce food waste effectively without complementary actions, such as menu planning and staff training. The study’s reliance on qualitative data and prototype simulations limits its generalizability, and restaurants may need context-specific data to accurately quantify the impact of interventions. It is estimated that 17% of food is wasted at the retail level, where redistribution can have a bigger impact than at the production or consumption stages of the supply chain. Although food waste at the retail level has been studied less than at other supply chain stages, retail stores tend to generate more food waste than food loss, as the discarded food is generally still consumable.

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(usmap)
library(ggplot2)
library(scales)
library(Metrics) 
library(caret)
library(randomForest)
library(stringr)
library(jsonlite)
library(httr)
library(purrr)
```

## Data

Food loss and waste data is scarce due to the uneven information across all stages of food chains, and aggregated data that is difficult to replicate. In fact, a UNEP study [@SolohaDace2025] noted that the food waste per capita in the EU cannot be fully explained because the methodology used to quantify food waste in each member state is not known, making it difficult to conclude whether the differences in data are justified or are due to differences in scope. One review paper, [@xue2017missing]. identified that data is heavily biased towards industrialized and high-income countries. The paper also points out that fewer studies focus on retail, households, and distribution levels of the supply chain.

## Food insecurity

The food environment atlas [@USDA_Food_Environment_Atlas] provides secondary data by state and county. The data set includes the food insecurity, demographics, and food-program participation rates. For our model we will include the makeup of the counties race, the median income, obesity rate, and food programs participation rate changes over a period of years.

    Foodinsecurity - household food insecurity

    WhitePct - White Race percantage

    BlackPct - Black Race percantage

    AsianPct - Asian Race percentage

    NativePct - Native Race percentage

    PacificIslanderPct - Pacific Race percentage

    HispPct - Hispanic Race percentage

    TwoOrMorePct - Two Race percentage

    Medianincome - Median ncome

    NocarlowaccessChange - No car low acces to stores changes

    ObesityPct   Obesity Rate percentage

    WICChange - WICC participation change

    SNAPparticipantschange - SNAP participation change

    SNAPstoreschang - SNAP store participation change

    SummerFoodProgChange - Summer food  program rate change

    MetroStatus - Metro  avilability

    adultdiabetesrate - Adults diabetes rate

    WomenWICpart - Women WIC participation

    Excessfoodrestaurant - Excess food Restaurant estimates

## Restaurant excess estimates

US EPA [@EPA_ExcessFoodOpportunitiesMap] provides a map of restaurants at the county level with excess food estimates. We will use this data to find restaurant average food loss at the county level; we will then combine this data with the food insecurity data set.

Avg_Excess_Food -Average restaurant Excess food in tons

# Data Exploration

The average food insecurity rate from 2021 to 2023 is mostly evenly distributed with the mean for counties across the United States at 12.46% and median of 12.30%. The race demographics show, as expected, more counties with high non-Hispanic white populations, while minorities are right skewed. households with no car and low access to scores is roughly normal with some counties experiencing low access to stores without cars. The median household income in the united states is about \$56K; however, some counties have a median income over \$150K.

The map graph identifies at the state levels that high levels of food insecurity is prevalent in the south. For example Texas, Arkansas, and Alabama have high levels of food insecurity. Another graph shows that New Mexico has high food waste estimates from restaurants. Finally, states like Nevada, New mexico, South Carolina, Michigan, and Ohio have high rates of food insecurity as well as food excess from res taunts.

Higher participation of women in WIC programs correlates with higher food insecurity. Notably, the percentage change in WIC participants between 2017 and 2021 has a strong negative correlation with food insecurity (-52%). Since the mean percentage change of WIC participants is -30%, we can conclude that if participation rates increase, food insecurity is likely to decrease. Changes in SNAP participation are also negatively correlated with food insecurity, with decreases in SNAP participation associated with higher food insecurity.

The availability of fast food restaurants has the highest positive correlation (10.7%) with restaurant excess food estimates. In addition, WIC redemptions are also positively correlated with excess food estimates from restaurants. Finally, grocery store availability shows a weak negative correlation (–14%) with restaurant excess food estimates, indicating that areas with more grocery stores tend to have less excess food from restaurants.

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
insecurity <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/StateAndCountyData.csv")

Restaurant <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/RestaurantsAndFoodServices.csv")

pop <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/co-est2022-alldata.csv")

fi <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/2025%20County%20Health%20Rankings%20Data%20-%20v4%20(1).csv")
```

```{r}
# population 

state_lookup <- data.frame(
  StateFull = state.name,
  StateAbbrev = state.abb
)

pop <- pop %>%
  mutate( CTY = COUNTY ,  County = CTYNAME,  ST = STATE,  State = STNAME,  POP22 = POPESTIMATE2022, County = str_remove(County," County$"), County = str_trim(County), State = str_trim(State)
)%>%
    left_join(state_lookup, by = c("State" = "StateFull")) %>%
  mutate(State = StateAbbrev) %>%
  select(State,County, POP22)

glimpse(pop)
```

```{r}
pop <- pop %>%
  group_by(State, County) %>%
  filter(n() == 1) %>%  # keep only unique State+County combinations
  ungroup()

glimpse(pop)
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
insecurity_wider <- insecurity %>%
  pivot_wider(.,names_from = Variable_Code, values_from = Value)

#glimpse(insecurity_wider)
```

```{r}

insecurity_wider <-  insecurity_wider %>%
  mutate(PERPOV17_21 = as.factor(PERPOV17_21),METRO23= as.factor(METRO23), POPLOSS15 =as.factor(POPLOSS15) ) 
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
insecurity_wider <- insecurity_wider %>%
  
  mutate(across(where(is.numeric), ~replace(., . %in% c(-8888, -9999), NA))) %>%
  
  mutate(
    PERPOV17_21 = as.factor(replace(PERPOV17_21, PERPOV17_21 %in% c(-8888, -9999), NA)),
    METRO23 = as.factor(replace(METRO23, METRO23 %in% c(-8888, -9999), NA)),
    POPLOSS15 = as.factor(replace(POPLOSS15, POPLOSS15 %in% c(-8888, -9999), NA))
  )
```

```{r}
# combine population to insecurity data

insecurity_wider <- insecurity_wider %>%
  left_join(pop, by = c("State", "County"))

#glimpse(insecurity_wider)
```

```{r}
# calculate total cnap $'s per county, SNAP population, and SNAP$ per participant
insecurity_wider <- insecurity_wider %>% 
  mutate( total_SNAP = PC_SNAPBEN22*POP22,
          SNAP_participants = POP22* (PCT_SNAP22/100),
          SNAP_per_participant= (total_SNAP/SNAP_participants)
          )
#glimpse(insecurity_wider)
```

```{r}
# 2021 market basked adult = 2300 kcal/day *2 + 1600 kcal/day + 1800 kcal day  = # 8000 per month / 835.57 = 316 per dollar
# add calories per month

insecurity_wider <- insecurity_wider%>%
  mutate(calories_per_person = SNAP_per_participant * 316)
 
```

```{r}
fi <- fi %>% 
  select(State, County, Food_insecure)
insecurity_wider <- insecurity_wider %>%
  left_join(fi, by = c("State", "County"))

glimpse(insecurity_wider)
```

```{r}
# use relevane variables
insecurity_wider <- insecurity_wider %>%
  select( State, County,FOODINSEC_21_23,PCT_NHWHITE20, PCT_NHBLACK20, PCT_NHASIAN20,PCH_WICWOMEN_16_21, PCT_NHNA20, PCT_HISP20, PCT_NHTMR20, MEDHHINC21, PCH_SNAP_17_22, PCH_LACCESS_LOWI_15_19, PCH_LACCESS_HHNV_15_19, PCT_OBESE_ADULTS22, PCH_SFSP_17_21, PCH_WIC_17_21, REDEMP_SNAPS23,PC_SNAPBEN22,REDEMP_WICS22,SNAPSPTH23, PCH_SNAPSPTH_17_23, PCT_LACCESS_SNAP19, POVRATE21, METRO23,total_SNAP, SNAP_participants , SNAP_per_participant, calories_per_person)

glimpse(insecurity_wider)


```

```{r}
numeric_vars <- names(insecurity_wider)[sapply(insecurity_wider, is.numeric)]

summary_table <- insecurity_wider %>%
  summarise(across(all_of(numeric_vars),
                   list(
                     mean = ~mean(., na.rm = TRUE),
                     median = ~median(., na.rm = TRUE),
                     sd = ~sd(., na.rm = TRUE),
                     min = ~min(., na.rm = TRUE),
                     max = ~max(., na.rm = TRUE),
                     n_missing = ~sum(is.na(.))
                   ),
                   .names = "{.col}_{.fn}")) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", "stat"),
    names_pattern = "^(.*)_(mean|median|sd|min|max|n_missing)$",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  arrange(variable)
summary_table <- summary_table %>%
  mutate(across(where(is.numeric), ~round(., 2)))

tail(summary_table)

print(numeric_vars)
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}

plots <- lapply(numeric_vars, function(col_name){
  ggplot(insecurity_wider, aes_string(x = col_name))+
    geom_histogram(fill = "red", bins = 20, na.rm = TRUE)+
    labs(title = col_name, x = col_name, y = "count")+
      theme_minimal()
})



plots

```

```{r}
scatter_plots <- lapply(numeric_vars, function(col_name){
  ggplot(insecurity_wider, aes_string(x = col_name, y = "FOODINSEC_21_23")) +
    geom_point(alpha = 0.6, color = "blue")+
    labs(title = paste("FOODINSEC_21_23 vs", col_name),
         x = col_name,
         y = "FOODINSEC_21_23") +
    theme_minimal()
})



scatter_plots
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
Restaurant <- Restaurant %>%
  mutate(Excess_food = 
           (`Excess.Food.Estimate..Low..tons.per.year.` +
              `Excess.Food.Estimate..High..tons.per.year.`) / 2) %>%
  group_by(State, County) %>% 
  summarise(
    Total_Excess_Food = sum(Excess_food, na.rm = TRUE),
    n_restaurants = n()  # optional, tells you how many restaurants per county

  ) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

glimpse(Restaurant)
```

```{r}
#combine data
combined<- insecurity_wider %>%
  left_join(Restaurant, by = c("State", "County"))

glimpse(combined)




```

Mismatch between food insecurity and food loss

```{r}
# food insecurity by state
food_insec_state <- combined %>%
  group_by(State) %>%
  summarize(Food_Insecurity = mean(FOODINSEC_21_23, na.rm = TRUE)) %>%
  ungroup() %>%
  rename(state = State) %>%
  mutate(state = toupper(state))

#average restaurant excess food
excess_food_state <- combined %>%
  group_by(State) %>%
  summarize(Avg_Excess_Food = mean(Avg_Excess_Food, na.rm = TRUE)) %>%
  ungroup() %>%
  rename(state = State) %>%
  mutate(state = toupper(state))

state_food <- left_join(food_insec_state, excess_food_state, by = "state")

# food insecurity by state
plot_usmap(data = state_food, values = "Food_Insecurity", regions = "states") +
  scale_fill_continuous(
    name = "Food Insecurity (21-23)",
    low = "lightblue",
    high = "darkblue",
    na.value = ""
  ) +
  labs(title = "Food Insecurity by State (2021-2023)")

# Averege restaurant excess food by state
plot_usmap(data = state_food, values = "Avg_Excess_Food", regions = "states") +
  scale_fill_continuous(
    name = "Avg Excess Food",
    low = "lightblue",
    high = "darkblue",
    na.value = "black"
  ) +
  labs(title = "Average Excess Food by State")
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}

# US food insecurity and Restaurant excess food hotspots
state_food <- state_food %>%
  mutate(
    Excess_Food_Avg = mean(Avg_Excess_Food, na.rm = TRUE),
    Hotspot = case_when(
      Food_Insecurity > 12.46 & Avg_Excess_Food > Excess_Food_Avg ~ "High Food Insecurity & High Excess Food",
      Food_Insecurity > 12.46 ~ "High Food Insecurity",
      Avg_Excess_Food > Excess_Food_Avg ~ "High Excess Food",
      TRUE ~ "Low"
    )
  )

plot_usmap(data = state_food, values = "Hotspot", regions = "states") +
  scale_fill_manual(
    values = c(
      "High Food Insecurity & High Excess Food" = "red",
      "High Food Insecurity" = "orange",
      "High Excess Food" = "yellow",
      "Low" = "lightblue"
    ),
    name = "State Category"
  ) +
  labs(title = "US Food Insecurity vs Excess Food Hotspots (2021-2023)") +
  theme(legend.position = "right")
```

```{r}
corr_with_foodinsec <- combined %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ cor(.x, combined$FOODINSEC_21_23,
                                       use = "pairwise.complete.obs"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))

tail(corr_with_foodinsec)
```

# Data Preparation

Attempt to normalize the variables that are not percent changes

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
combined <- combined %>%
  mutate(
    across(
      # choose numeric columns
      .cols = where(is.numeric) &
              # exclude dependent variable
              !matches("^FOODINSEC_21_23$") &
              # exclude variables beginning with "PCH"
              !matches("^PCH_"),
      # apply log1p safely to avoid issues with 0 values
      .fns = log1p
    )
  )

glimpse(combined)

```

```{r}
combined <- combined %>%
  group_by(State) %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()


combined <- combined %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Winsorize to remove outliers
winsorize <- function(x, probs = c(0.01, 0.99)) {
  q <- quantile(x, probs = probs, na.rm = TRUE)
  pmin(pmax(x, q[1]), q[2])
}

combined <- combined %>%
  mutate(across(where(is.numeric), winsorize))

# Convert factor to numeric
combined <- combined %>%
  mutate(
    METRO23 = as.numeric(as.character(METRO23)),  # convert factor levels to numeric
    METRO23 = as.numeric(mode_impute(METRO23))     # impute mode and ensure numeric
  )

na_counts <- colSums(is.na(combined))
na_columns <- names(na_counts[na_counts > 0])
na_columns

head(combined)
```

```{r}

check <- combined %>%
  select(where(~ any(is.na(.))))

summary(check)


```

```{r}

cat(names(combined), sep = "\n")

# Only numeric variables as a vertical list
numeric_vars <- names(combined)[sapply(combined, is.numeric)]
cat(numeric_vars, sep = "\n")

# Only factor/character variables as a vertical list
factor_vars <- names(combined)[sapply(combined, function(x) is.factor(x) | is.character(x))]
cat(factor_vars, sep = "\n")
```

```{r}
## rename columns for clarity
combined <- combined %>%
  rename(
    Foodinsecurity = FOODINSEC_21_23,
    WhitePct = PCT_NHWHITE20,
    BlackPct = PCT_NHBLACK20,
    AsianPct = PCT_NHASIAN20,
    NativePct = PCT_NHNA20,
    HispPct = PCT_HISP20,
    TwoOrMorePct = PCT_NHTMR20,
    Medianincome = MEDHHINC21,
    low_income_low_store_access = PCH_LACCESS_LOWI_15_19,
    no_car_and_low_store_access= PCH_LACCESS_HHNV_15_19,
    ObesityPct = PCT_OBESE_ADULTS22,
    WICChange = PCH_WIC_17_21,
    SNAPparticipantschange = PCH_SNAP_17_22,
    SNAPstoreschange = PCH_SNAPSPTH_17_23,
    SummerFoodProgChange = PCH_SFSP_17_21,
    MetroStatus = METRO23,
    adult_obese_rate = PCT_OBESE_ADULTS22,
    WomenWICpart = PCH_WICWOMEN_16_21,
    Poverty_rate = POVRATE21,
    WIC_redemptions = REDEMP_WICS22,
    SNAP_redemptions = REDEMP_SNAPS23,
    SNAP_per_capita = PC_SNAPBEN22,
    SNAP_auth_stores = SNAPSPTH23,
    SNAP_low_access_pct = PCT_LACCESS_SNAP19
  )

model_data <- combined %>%
  select(-State, -County)
 
glimpse(model_data)

```

```{r}
selected_vars <- colnames(model_data)
data_select <- model_data %>%
  select(all_of(selected_vars)) %>%
  select(where(is.numeric))

data_corr <- cor(data_select, use = "complete.obs")


corrplot::corrplot(data_corr, method = "color", tl.col = "black", tl.srt = 45)
head(corr_table, 20)
  
```

```{r}
# 1. Select numeric variables
numeric_vars <- model_data %>% select(where(is.numeric))

# 2. Compute correlations with Foodinsecurity
cor_with_food <- sapply(numeric_vars, function(x) cor(x, numeric_vars$Foodinsecurity, use = "complete.obs"))

# 3. Convert to data frame
cor_df <- data.frame(
  Variable = names(cor_with_food),
  Correlation = round(cor_with_food, 2)
) %>%
  filter(Variable != "Foodinsecurity") %>%   # remove self-correlation
  arrange(desc(abs(Correlation)))

# 4. View table
cor_df

# 5. Plot correlations
ggplot(cor_df, aes(x = reorder(Variable, Correlation), y = Correlation)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Correlation with Foodinsecurity",
       x = "Variable",
       y = "Correlation") +
  theme_minimal()
```

```{r}
data_plot <- lapply(names(data_select), function(col_name){
  ggplot(data_select, aes_string(x =col_name)) +
    geom_histogram(fill = "red", bins = 20, na.rm = TRUE) +
    labs(title = col_name, x = col_name, y = "count") +
    theme_minimal()    
})

data_plot


```

# Models

## Methodology

This study models county level food insecurity based on demographics, economic factors, health, and government program participation. Prior research utilizes demographic statistics such as race and genders to understand the ethnic groups that may contribute the unequal access to food. This is especially true in the United States as the racial demographic of the United states is very diversified. undefined

Government assistance indicators are also used as predictors. Changes in participation in SNAP, WIC, the Summer Food Program, and other support programs provide insight into how resource availability and policy implementation may influence local food access. These variables serve as proxies for demand for assistance and the adequacy of food support infrastructure.

Socioeconomic conditions are represented by median household income, which is a strong determinant of food affordability and financial stability. Finally, county-level obesity rates are included because diet-related health outcomes often correlate with limited access to healthy and affordable foods. Together, these variables form a comprehensive predictive model for understanding the drivers of food insecurity across U.S. counties.

## linear model

```{r}

set.seed(1234)


sample<- sample(nrow(model_data), nrow(model_data) *.60,
                replace = FALSE)

```

```{r}
foodinsec_train <- model_data[sample,]
foodinsec_train

foodinsec_test <-  model_data[-sample,]
foodinsec_test
```

```{r}
# modeling all predictors
model1 <- lm(Foodinsecurity ~ ., data = foodinsec_train )
summary(model1)
```

```{r}



predictions <- predict(model1, newdata = foodinsec_test)

rmse_test <- rmse(foodinsec_test$Foodinsecurity, predictions)

rmse_test


```

```{r}
# model 2
set.seed(12345)
lm_model2 <- lm(
  Foodinsecurity ~ BlackPct + AsianPct + WomenWICpart + NativePct + HispPct + TwoOrMorePct +
                   Medianincome + low_income_low_store_access + adult_obese_rate + 
                   SummerFoodProgChange + WICChange + SNAP_redemptions + SNAP_per_capita +
                   WIC_redemptions + SNAP_auth_stores + SNAPstoreschange +
                    calories_per_person,
  data = foodinsec_train
)
summary(lm_model2)
```

```{r}
predictions2 <- predict(lm_model2, newdata = foodinsec_test)

rmse_test2 <- rmse(foodinsec_test$Foodinsecurity, predictions)

rmse_test2
```

```{r}
AIC(model1, lm_model2)
BIC(model1, lm_model2)
```

## Random Forest

```{r}
set.seed(111)
rf_model <- randomForest(
  Foodinsecurity ~ WhitePct + BlackPct + NativePct + 
    PacificIslanderPct +  Medianincome + NocarlowaccessChange + 
    ObesityPct + WICChange + SNAPparticipantschange + 
    SNAPstoreschang + SummerFoodProgChange + MetroStatus + 
    WomenWICpart,
  data = foodinsec_train,
  ntree = 500,
  importance = TRUE
)

print(rf_model)
importance(rf_model)
varImpPlot(rf_model)
```

```{r}
set.seed(1235)
control <- trainControl(method = "cv", number = 5)
rf_cv <- train(
  Foodinsecurity ~ WhitePct + BlackPct + NativePct + PacificIslanderPct +
    Medianincome + NocarlowaccessChange + ObesityPct + WICChange +
    SNAPparticipantschange + SNAPstoreschang + SummerFoodProgChange +
    MetroStatus + WomenWICpart,
  data = model_data,
  method = "rf",
  trControl = control,
  tuneLength = 3
)
print(rf_cv)
```

# Conclusion

The first model developed to predict food insecurity by county is a multiple linear regression which included key demographic variables, as well as economic, access to stores, and participation in food programs. This model scored an adjusted R\^2 of 71%. Most variables were statistically significant, except for the Asian population in the county and the adult obesity rate. As expected, median income, various types of food programs participation changes, and overall obesity rates were statistically significant. This indicates that some of the variables such as adult obesity rate was redundant and some multicollinearity.

The second model removed the weaker predictors, due to the insignificance to the model. However, the adjusted R\^2 remained about the same at \~71%. This concluded that the excluded variables were in fact not valuable to the model.

Lastly, the random forest model used the same predictors as the previous model to capture any nonlinear relationships among the variables. This modeled outperformed with an R\^2 of 99.97%, an almost perfect accuracy rate. WIC participation from women was the most influential variable, which is as expected because this variable had a high correlation with food insecurity.

```{r}
model_summary <- data.frame(
  Model = c("Linear Regression (Full)", 
            "Linear Regression (Reduced)", 
            "Random Forest"),
  RMSE = c(1.295, 1.29, 0.0385),
  R2 = c(0.7091, 0.7091, 0.99976),
  MAE = c(NA, NA, 0.0105)  
)

model_summary

```

```{r}
model_long <- model_summary %>%
  pivot_longer(cols = c(RMSE, R2), names_to = "Metric", values_to = "Value")


ggplot(model_long, aes(x = Model, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("R2" = "#1b9e77", "RMSE" = "#d95f02")) +
  labs(title = "Model Performance Comparison",
       x = "Model",
       y = "Metric Value") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
ggplot(model_long, aes(x = reorder(Model, -Value), y = Value, fill = Metric)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(title = "Model Comparison: Accuracy and Error",
       x = "Model", y = "Value") +
  theme_minimal(base_size = 13)
```
