---
title: "Food insecurity Intervention"
author: "Angel Gallardo"
date: "2025-11-08"
bibliography: library.bib
format: pdf
---

# Abstract

Food insecurity remains a global issue. In the United States alone, 13.5% of households, or 18 million households, were food insecure in 2023. At the same time, restaurants contribute to significant food loss, food that is ready for consumption. As such, restaurant food loss represents a potential but underutilized resource for intervention. This research aims to build a model that accurately predicts high food insecurity at the county level. Furthermore, we look to demonstrate the potential impact that local restaurants and government programs can create with efficient intervention. This way, solving food insecurity in the United States can also reduce food waste from restaurants in the community. Policy-making can improve at the county level to help mitigate food insecurity and food loss with food assistance programs, as opposed to the current state-level decision-making. By incorporating food assistance program participation data into the model, we can gain insight into how Americans are securing their food. Thus, this research looks to detect whether calorie intake can predict high food insecurity. Finally, this research simulates an intervention, where a portion of restaurant food loss is converted into recovery calories through SNAP policy changes.

# Introduction

In 2023, the USDA, FDA, and EPA [@fda2023nationalstrategy]announced a national goal to reduce food loss and waste in the U.S by 50% by 2030. Food insecurity [@feedingamerica_foodinsecurity] is described as people not having access to food needed to live their fullest lives. On the other side, Food waste [@ReFED2025] is defined as the edible amount of food, post harvest, that is available for human consumption but is not yet consumed for any reason. Currently, more tha one-third of all available food goes uneaten through loss or waste.

According to Feeding America [@feedingamerica_foodinsecurity], the primary drivers of food insecurity include income-related factors (low-wage jobs, unreliable work, or financial emergencies); high cost of living (essentials are increasingly difficult to afford); community factors (limited access to transportation); health-related issues (reduced budgets for food and other essentials); and systematic barriers (unfair systems in policies and institutions).

By using readily available secondary data such as income levels, community factors, health issues, and combining it with government assistance program participation we can:

Using readily available secondary data—including income levels, community characteristics, health indicators, and participation in government food assistance programs—this research seeks to (1) build a model that identifies counties most in need of intervention and predicts high food insecurity, and (2) determine the demographic and program participation variables most strongly associated with food insecurity in the United States. Finally, the study simulates a policy-driven intervention in which a portion of restaurant food loss is recovered and redirected through SNAP, estimating its potential impact on reducing food insecurity at the county level.

# Research Question

This research focuses on county level, where a model can reveal high priority counties most in need of food security intervention. By developing a model capable of predicting high priority counties, government programs can rapidly focus on collaborating with local retailers to mitigate hunger. In addition, local retailers may also benefit by reduction of food loss and potential revenue increase when participating in discounted food assistance programs. In essence, this model could be utilize for state leaders to collaborate with local retailers, or the government can use this for policy making to increase restaurant food program authorization where needed. Currently, each state sets their own requirements for choosing how many and which restaurants are allowed to participate in the SNAP program. A county-level modeling approach can support state leaders in expanding authorized food program retailers where they are most needed, strengthening local food systems and improving program accessibility.

The significance of this phenomenon is twofold. First, more than 95 percent of food waste in the United States ends up in landfills, where it breaks down anaerobically and releases methane, carbon dioxide, and other greenhouse gases into the atmosphere [@mu2017environmental]. If food waste is not addressed, it will continue to contribute significantly to climate change. Second, the USDA reported that in 2023, 13.5 percent of U.S. households experienced food insecurity at some point during the year. Food insecurity is a global crisis, and countries around the world are actively seeking solutions to eliminate it.

Addressing food waste and insecurity has numerous real-world applications. For businesses, especially in agriculture, food production, and retail-reducing food waste can lead to cost savings, more efficient supply chains, and improved sustainability practices. Food that would otherwise go to waste can be redistributed to food banks and local communities, helping to reduce hunger in America and improve public health outcomes. Local non-profits and government can use this model to develop targeted programs and initiatives that connect surplus food with those in need. Taxpayers also benefit when program benefits is spent efficiently. Furthermore, reducing food waste directly contributes to the fight against climate change by lowering greenhouse gas emissions and easing the strain on natural resources. In the long term, tackling this issue supports environmental sustainability, economic efficiency, and social equity.

**Research Questions**

-   Can a machine learning model accurately identify counties with high food insecurity in the United States?

-   Which demographic, socioeconomic, and food assistance indicators are most significant in predicting food insecurity?

-   Can a policy intervention that recovers a portion of restaurant food loss reduce the number of high food insecurity counties?

I hypothesize that food recovery from restaurants may help reduce food insecurity in areas with low store accessibility.

# Literature Review

Measuring Food insecurity and Hunger in the United States [@carlson1999measuring] developed measuring scale to quantify food insecurity in the US. The measurement scale included household surveys where the questions were categorically measured to reflect the severity of the responses. This benchmark classifies food security to be defined as assured access at all times to enough food for an active healthy life. The other form of food insecurity, malnutrition, is not the case in the US, as there are no clinical symptoms of deprivation. As such, questions from skipping meals or children being hungry estimated a higher severity of food insecurity. The results show that severe hunger was more prevalent in households with children under eighteen years old. Additionally, the scale identified that food insecurity was more common among minority households and low income to poverty household ratios. This scale serves as a baseline to understand the relationship between food insecurity and key demographics. However, a predictive model can utilize these predictors, as well as economic access, and calorie intake data to estimate the probability that a household becomes food insecured. In addition, the estimated recoverable

One research paper [@lentz2019data] pointed out that the Integrated Food Security Phase Classification (IPC), a global framework that classifies food insecurity, does not make use of the readily available secondary data, in a replicable and transparent manner. Furthermore, there is no current food security early warning and monitoring system that incorporates readily available data into a predictive model. As a result, these methods limit timely, targeted humanitarian response. The research builds linear and log-linear models using three readily available secondary data measures, weather and price fluctuations, geographic variations, demographics, and intra-annual seasonal trends to predict food security at the sub national level. Their findings show that their model correctly predicts 83% to 99% of the most food insecure clusters, while the IPC’s approach only predicts between 0% and 10%. Although the model is an improvement from the IPC's approach, the model does not incorporate solutions to reduce food insecurity. Meanwhile, the U.S, unlike other countries, weather changes or geographic variations may not be useful as food insecurity is mainly driven by food program access, unemployment, and household characteristics. In addition, by accurately predicting high food insecurity, the US government can assess whether counties are steering towards high food insecurity to perform early intervention.

Leveraging Machine Learning for Food Waste Reduction: An Analysis of Predictive Models [@yang2024foodwaste] predicted food waste using data on household consumption patterns, retail demand, and food service estimates; along with confidence in estimates to gauge reliability for each country. It focuses on machine learning algorithms like Random Forest to predict food waste patterns for support policymakers in developing policies that facilitate interventions to reduce food loss. The report concluded that household-level food waste estimates are the most influential factors driving food waste predictions at the per-capita level. However, household data may be hard to collect due to the reliability of the self-reported data. In addition, collecting reliable household data remains challenging due to the limitations of self-reported information and the lack of incentives for households to participate in surveys. As a result, using household food excess as a solution may be less practical or effective to redistribute excess food.

Another study [@irani2018managing] examines how food waste and loss in supply chains impact food security, using causal modeling and simulation to identify key drivers from qualitative data. The findings show that food waste is not solely the result of kitchen inefficiencies or over-ordering, but arises from a combination of factors, including consumer behavior, storage logistics, and supplier practices. However, one study was noted as identifying plate waste to compomise the highest amount of food waste, followed by storage, and and preparation loss. The research employs a Fuzzy Cognitive Map (FCM) to model the interrelated factors driving food waste and loss. The study concludes that no single factor—such as inventory management—can reduce food waste effectively without complementary actions, such as menu planning and staff training. The study’s reliance on qualitative data and prototype simulations limits its generalizability, and restaurants may need context-specific data to accurately quantify the impact of interventions. It is estimated that 17% of food is wasted at the retail level, where redistribution can have a bigger impact than at the production or consumption stages of the supply chain. Although food waste at the retail level has been studied less than at other supply chain stages, retail stores tend to generate more food waste than food loss, as the discarded food is generally still consumable.

Currently, the Restaurant Meal Program (RMP) allows certain SNAP clients to buy prepared meals from restaurants using their SNAP benefits. This program is available in a handful of states; to qualify, a client must be SNAP verified, elderly, disabled, homeless, or a spouse of a qualifying client for the RMP program. The goal for this research is not only to predict counties with high food insecurity, but if and where the program extension is the most impact. Overall, solving food insecurity benefits everyone, from residents, food retailers, and taxpayers.

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(usmap)
library(ggplot2)
library(scales)
library(Metrics) 
library(caret)
library(randomForest)
library(stringr)
library(jsonlite)
library(httr)
library(purrr)
library(skimr)
library(GGally)
library(ggforce)
library(corrplot)
```

## Data

## Food insecurity

The food environment atlas [@USDA_Food_Environment_Atlas] provides secondary data by state and county. The data set includes the food insecurity by state only, demographics, and food-program participation rates. For our model we will include the makeup of the counties race and age, the median income, obesity rate, and food programs participation rates.

The re-evaluation of the Thrifty food Plan [@USDA_ThriftyFoodPlan2021] in 2021 to estimate the cost of a practical, nutritious, and budget-conscious diet. In 2021 the food plan referenced a family of four, where the cost of the market basket was calculated at \$835.57 per month. The referenced family consisted of two adults, male and female, between the ages of 20-50 years old, calculated a daily intake of calories of about 3,000 and 2,200. The other family members included two children, one with the age rage 6-8, and a daily calorie intake of 1,800, and the other in the 9-11 age category with a daily intake of 2,200. This means that on average one dollar of SNAP benefit purchases 330 calories. This information will be utilized to create additional variable- calories per SNAP recipient- where we can visualize whether the adequate allotment is given to participants.

## Restaurant excess estimates

Food loss and waste data is scarce due to the uneven information across all stages of food chains, and aggregated data that is difficult to replicate. In fact, a UNEP study [@SolohaDace2025] noted that the food waste per capita in the EU cannot be fully explained because the methodology used to quantify food waste in each member state is not known, making it difficult to conclude whether the differences in data are justified or are due to differences in scope. One review paper, [@xue2017missing] identified that data is heavily biased towards industrialized and high-income countries. The paper also points out that fewer studies focus on retail, households, and distribution levels of the supply chain.

US EPA [@EPA_ExcessFoodOpportunitiesMap] provides a map of restaurants at the county level with excess food estimates. We will use this data to find the aggregate food loss in tons from restaurants at the county level. This data will be utilized to estimate the amount of available calories that can be recoverable at the local level. Finally, a percentage of the recoverable amount will be estimated and simulated to calculate the impact it has to reduce food insecurity.

```{r}
insecurity <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/StateAndCountyData.csv")

fi <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/2023%20County%20Health%20Rankings%20Data%20-%20v2.csv")

Restaurant <- read.csv("https://raw.githubusercontent.com/Angelogallardo05/data698/refs/heads/main/RestaurantsAndFoodServices.csv")

```

```{r}
glimpse(fi)
```

```{r}
glimpse(insecurity_wider)
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
insecurity_wider <- insecurity %>%
  pivot_wider(.,names_from = Variable_Code, values_from = Value)

#glimpse(insecurity_wider)
```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
insecurity_wider <- insecurity_wider %>%
  
  mutate(across(where(is.numeric), ~replace(., . %in% c(-8888, -9999), NA))) %>%
  
  mutate(
    PERPOV17_21 = as.factor(replace(PERPOV17_21, PERPOV17_21 %in% c(-8888, -9999), NA)),
    METRO23 = as.factor(replace(METRO23, METRO23 %in% c(-8888, -9999), NA)),
    POPLOSS15 = as.factor(replace(POPLOSS15, POPLOSS15 %in% c(-8888, -9999), NA))
  )
```

```{r}
state_lookup <- data.frame(
  State = state.name,
  StateAbbrev = state.abb,
  stringsAsFactors = FALSE
)

fi <- fi %>%
  left_join(state_lookup, by = c("State" = "State")) %>%
  mutate(State = StateAbbrev) %>%       # replace with abbreviations
  select(-StateAbbrev)

fi <- fi %>% 
  select(State, County, Food_insecure = X..Food.Insecure.1  ,Population, Household_burden = X..Households.with.Severe.Cost.Burden.1, Rural_percent =X..Rural  )
insecurity_wider <- insecurity_wider %>%
  left_join(fi, by = c("State", "County"))

glimpse(insecurity_wider)
```

additional SNAP variables

```{r}
insecurity_wider <- insecurity_wider %>% 
  mutate( total_SNAP = PC_SNAPBEN22*Population,
          SNAP_participants = Population* (PCT_SNAP22/100),
          SNAP_per_participant= (total_SNAP/SNAP_participants)
          )
#glimpse(insecurity_wider)   
```

add calories per month

```{r}

# add calories per month

insecurity_wider <- insecurity_wider%>%
  mutate(calories_SNAP_part = SNAP_per_participant * 330)
```

```{r}

```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
Restaurant <- Restaurant %>%
  mutate(Excess_food = 
           (`Excess.Food.Estimate..Low..tons.per.year.` +
              `Excess.Food.Estimate..High..tons.per.year.`) / 2) %>%
  group_by(State, County) %>% 
  summarise(
    Total_Excess_Food = sum(Excess_food, na.rm = TRUE),
    n_restaurants = n()  

  ) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

glimpse(Restaurant)
```

```{r}
#combine data
combined<- insecurity_wider %>%
  left_join(Restaurant, by = c("State", "County"))




#Estimated calories wasted per capita
combined <- combined %>%
  mutate(kcal_wasted_per_capita = (Total_Excess_Food*2000*1400)/Population,
         restaurants_per_capita = n_restaurants/Population)

glimpse(combined)

```

```{r}
# use relevane variables
combined <- combined %>%
  select( State, County,Food_insecure ,PCT_NHWHITE20, PCT_NHBLACK20, PCH_WICWOMEN_16_21, PCT_HISP20, PCT_NHTMR20, MEDHHINC21, PCH_SNAP_17_22, PCH_LACCESS_LOWI_15_19, PCH_LACCESS_HHNV_15_19, PCT_OBESE_ADULTS22, PCH_SFSP_17_21, PCH_WIC_17_21, REDEMP_SNAPS23,PC_SNAPBEN22,REDEMP_WICS22,SNAPSPTH23, PCT_LACCESS_HHNV19, PCH_SNAPSPTH_17_23, PCT_LACCESS_SNAP19, POVRATE21, METRO23,total_SNAP, SNAP_participants , SNAP_per_participant, calories_SNAP_part ,PCT_65OLDER20, PCT_18YOUNGER20,Population, Household_burden, Rural_percent,Total_Excess_Food , n_restaurants, kcal_wasted_per_capita, restaurants_per_capita
 )

glimpse(combined)
```

# Data Exploration

The data summary shows that most of the variables data is complete, except for the WIC redemption variable at 55% complete. PCH_LACCESS_LOWI_15_19, defined as low income and low access to stores between 2015-2019 has a large standard deviation which could be due to population migration. The mean food insecurity rate per county is at 12.5%, with a standard deviation of 3%. The highest percent of food insecurity is at 29%. As expected, the white population per county is right-skewed, meaning that most counties has high white population. The mean poverty rate in the US is at 14.5%, where some counties experience poverty rates as high as 44%.

```{r}
options(scipen = 999)   

skim(combined)
```

The histogram plots demonstrate that most of the variables are either left or right skewed. For instance, the the histogram shows most counties have a high percentage of white residents, but there are extreme cases where counties have a low white population. However, food insecurity, poverty, population of residents under the age of 18, and over 65 are normally distributed. Because the dependent variable, food insecurity, is approximately normally distributed, it is more compatible with linear regression assumptions and may improve model accuracy. The skewed independent variables show that a transformation or a model robust to non-normal predictors may be more appropriate. In addition, the scatter plots confirm that most predictors exhibit heteroscedasticity against food insecurity, meaning that the variance is not consistent across the range of predictors. Overall, the the skewed predictors, as well as the presence of outliers indicate that variable transformation or non-linear models are more adequate to predict food insecurity.

```{r}
numeric_vars<- names(combined)[sapply(combined, is.numeric)]
combined_long <- combined %>%
  pivot_longer(cols = all_of(numeric_vars),
               names_to = "variable",
               values_to = "value")

ggplot(combined_long, aes(x = value)) +
  geom_histogram(bins = 20) +
  facet_wrap_paginate(~ variable, scales = "free", ncol = 5, nrow = 5, page = 1)
```

```{r}

numeric_vars <- setdiff(names(combined)[sapply(combined, is.numeric)], "Food_insecure")

combined_long <- combined %>%
  pivot_longer(cols = all_of(numeric_vars),
               names_to = "variable",
               values_to = "value")

ggplot(combined_long, aes(x = value, y = Food_insecure)) +
  geom_point(alpha = 0.5, color = "blue") +
  facet_wrap_paginate(~ variable, scales = "free_x", ncol = 5, nrow = 4, page = 1) +
  theme_minimal() +
  labs(x = "Variable Value", y = "Food Insecure",
       title = "Predictors vs Food Insecure")

```

The top predictors high positive correlation with food insecurity are poverty, SNAP participants that have low access to stores, and calories per SNAP dollar. The only variable highly negative corrlated with food insecurity, median income, demonstrates that higher income reduces food insecurity.

```{r}
corr_with_foodinsec <- combined %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ cor(.x, combined$Food_insecure,
                                       use = "pairwise.complete.obs"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))

corr_with_foodinsec
```

# Data Preparation

The NA's will be replaced by using the median of the state. Furthermore, in an attempt to normalize the predictors, large outliers are removed. If the predictors are not normalized, logistic models will be utilized to predict food insecurity. The dependent variable would be converted as a binary variable, where any value higher than the median (12%) would be classified as high food insecurity.

Certain variables will be renamed for clarity.

An additional variable, recovered calories from restaurant, will be creating using an example scenario, where SNAP allots additional dollars to purchase meals from restaurants at a discount rate. The model will used the same discount rate used in the restaurant meals program.

```{r}
mode_impute <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

combined <- combined %>%
  group_by(State) %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()


combined <- combined %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Winsorize to remove outliers
winsorize <- function(x, probs = c(0.01, 0.99)) {
  q <- quantile(x, probs = probs, na.rm = TRUE)
  pmin(pmax(x, q[1]), q[2])
}

combined <- combined %>%
  mutate(across(where(is.numeric), winsorize))


combined <- combined %>%
  mutate(
    METRO23 = as.numeric(as.character(METRO23)),  
    METRO23 = as.numeric(mode_impute(METRO23))     
  )

na_counts <- colSums(is.na(combined))
na_columns <- names(na_counts[na_counts > 0])
na_columns

head(combined)

```

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
corr_with_foodinsec2 <- combined %>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ cor(.x, combined$Food_insecure,
                                       use = "pairwise.complete.obs"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))

corr_with_foodinsec2


top_vars <- corr_with_foodinsec2 %>%
  slice_max(order_by = abs(Correlation), n = 15) %>%
  pull(Variable)
top_vars <- setdiff(top_vars, "Food_insecure")

top_data <- combined %>%
  select(all_of(top_vars))


ggpairs(top_data,
        upper = list(continuous = wrap("cor", size = 2)),
        lower = list(continuous = wrap("points", alpha = 0.5, size = 1.5)),
        diag = list(continuous = wrap("barDiag", fill = "lightblue"))
) + 
theme(strip.text = element_text(face = "bold",size = 7,  angle = 60))

```

```{r}
combined <- combined %>%
  mutate(High_FoodInsecure = ifelse(Food_insecure >= 12, 1, 0))

```

```{r}
Restaurant_snap_share <- 0.05
Res_price_per_cal <- 0.025
discount_rate <- 0.10
effective_price <- Res_price_per_cal * (1 - discount_rate)
recoverable_cal_rate <- 0.15  # adjustable

combined <- combined %>%
  mutate(
    SNAP_restaurant_spend = total_SNAP * Restaurant_snap_share,
    calories_snap_can_buy_rest = SNAP_restaurant_spend / effective_price,
    calories_wasted = kcal_wasted_per_capita * Population,
    recoverable_calories = calories_wasted * recoverable_cal_rate,
    Recovered_Calories = pmin(recoverable_calories, calories_snap_can_buy_rest)
  )

```

```{r}

check <- combined %>%
  select(where(~ any(is.na(.))))

summary(check)


```

```{r}

cat(names(combined), sep = "\n")

# Only numeric variables as a vertical list
numeric_vars <- names(combined)[sapply(combined, is.numeric)]
cat(numeric_vars, sep = "\n")

# Only factor/character variables as a vertical list
factor_vars <- names(combined)[sapply(combined, function(x) is.factor(x) | is.character(x))]
#cat(factor_vars, sep = "\n")
```

```{r}
## rename columns for clarity
combined <- combined %>%
  rename(
    Foodinsecurity = Food_insecure,
    WhitePct = PCT_NHWHITE20,
    BlackPct = PCT_NHBLACK20,
    HispPct = PCT_HISP20,
    TwoOrMorePct = PCT_NHTMR20,
    Medianincome = MEDHHINC21,
    low_income_low_store_access = PCH_LACCESS_LOWI_15_19,
    ObesityPct = PCT_OBESE_ADULTS22,
    WICChange = PCH_WIC_17_21,
    SNAPparticipantschange = PCH_SNAP_17_22,
    SNAPstoreschange = PCH_SNAPSPTH_17_23,
    SummerFoodProgChange = PCH_SFSP_17_21,
    MetroStatus = METRO23,
    adult_obese_rate = PCT_OBESE_ADULTS22,
    WomenWICpart = PCH_WICWOMEN_16_21,
    Poverty_rate = POVRATE21,
    WIC_redemptions = REDEMP_WICS22,
    SNAP_redemptions = REDEMP_SNAPS23,
    SNAP_per_capita = PC_SNAPBEN22,
    SNAP_auth_stores = SNAPSPTH23,
    SNAP_low_access_pct = PCT_LACCESS_SNAP19,
    Pct_65_older = PCT_65OLDER20,
    PCT_18_younger = PCT_18YOUNGER20,
    No_car_low_store = PCH_LACCESS_LOWI_15_19 
  )

model_data <- combined %>%
  select(-State, -County,-Foodinsecurity )
 
glimpse(model_data)

```

```{r}
corr_with_foodinsec2 <- model_data%>%
  select(where(is.numeric)) %>%
  summarise(across(everything(), ~ cor(.x, combined$High_FoodInsecure,
                                       use = "pairwise.complete.obs"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  arrange(desc(abs(Correlation)))

corr_with_foodinsec2
  
```

# Models

## Methodology

Due to persistant feature of skewness and heteroscedastic behavior against food insecurity, the model of choice is the logistic model. This models will identify the probability of high food insecurity by county demographics, economic factors, health, and government program participation. Although some demographics had low correlation against food insecurity, the first model will include all age and race predictor because prior research included those statistics to classify those groups that may contribute the unequal access to food. This may be especially true in the United States as the racial demographic of the United states is very diversified.\
\
Government assistance participation predictors such as participation in SNAP, WIC, the Summer Food Program, and other support programs are included in the first model to provide additional insight on how resource availability and policy implementation may influence local food access. These variables serve as proxies for demand of assistance and the adequacy of food support infrastructure.\
\
Socioeconomic conditions are represented by median household income, which is a strong determinant of food affordability and financial stability. Household burden, or food resource limitations due to other financial obligations is also included because affordable foods does affect at least one member in the family , according to previous research. Together, these variables form a comprehensive predictive model for understanding the drivers of food insecurity across U.S. counties. The other variable included in the first model is the number of restaurants per county, with the hypothesize that access to restaurants will be a significant factor in predicting high food insecurity.\
\
\
Finally, the last model will include a potential calories that could be saved through SNAP and restaurant meal intervention. The recovered calories is the amount of restaurant meals that additional SNAP funds can recover if it was sold at a discount, as implemented in some areas. 

## linear model

```{r}

set.seed(1234)


sample<- sample(nrow(model_data), nrow(model_data) *.60,
                replace = FALSE)

```

```{r}
foodinsec_train <- model_data[sample,]
foodinsec_train

foodinsec_test <-  model_data[-sample,]
foodinsec_test
```

```{r}
# modeling all predictors
model1 <- glm(
  High_FoodInsecure ~ Poverty_rate + Medianincome + calories_SNAP_part+ WhitePct + TwoOrMorePct + BlackPct + HispPct + SNAP_low_access_pct + Rural_percent +  Pct_65_older + PCT_18_younger +restaurants_per_capita + Household_burden
 ,
  data = foodinsec_train,
  family = binomial
)
summary(model1)
```

```{r}
pred_probs <- predict(model2, newdata = foodinsec_test, type = "response")

# ---------------------------
#   2. Convert to Class Labels
# ---------------------------
threshold <- 0.5  # You can adjust this later
pred_class <- ifelse(pred_probs > threshold, 1, 0)

# ---------------------------
#   3. Accuracy
# ---------------------------
accuracy <- mean(pred_class == foodinsec_test$High_FoodInsecure)
print(paste("Accuracy:", round(accuracy, 4)))

# ---------------------------
#   4. Confusion Matrix
# ---------------------------
conf_matrix <- table(
  Actual = foodinsec_test$High_FoodInsecure,
  Predicted = pred_class
)
print("Confusion Matrix:")
print(conf_matrix)

# ---------------------------
#   5. AUC & ROC Curve
# ---------------------------
library(pROC)

roc_obj <- roc(foodinsec_test$High_FoodInsecure, pred_probs)
auc_val <- auc(roc_obj)
print(paste("AUC:", round(auc_val, 4)))

# Plot ROC Curve
plot(roc_obj, col = "blue", lwd = 2,
     main = "ROC Curve for Model2")

# ---------------------------
#   6. Optional: RMSE using probabilities
#       (Not ideal for logistic regression but allowed)
# ---------------------------
rmse_test <- sqrt(mean((foodinsec_test$High_FoodInsecure - pred_probs)^2))
print(paste("RMSE:", round(rmse_test, 4)))

# ---------------------------
#   7. Pseudo R-squared (McFadden's)
# ---------------------------
library(pscl)

pseudo_r2 <- pR2(model2)
print("Pseudo R2:")
print(pseudo_r2)

# ---------------------------
#   8. Optional: Compare Model1 and Model2
# ---------------------------
# Model1 must already exist
if (exists("model1")) {
  aic_m1 <- AIC(model1)
  aic_m2 <- AIC(model2)
  
  print(paste("AIC Model1:", aic_m1))
  print(paste("AIC Model2:", aic_m2))
  
  if (aic_m2 < aic_m1) {
    print("Model2 improves model fit (lower AIC).")
  } else {
    print("Model1 has better or similar fit.")
  }
}

```

```{r}
model02 <- glm(
  High_FoodInsecure ~ Poverty_rate + Medianincome + calories_SNAP_part +
    WhitePct + TwoOrMorePct + BlackPct + HispPct + SNAP_low_access_pct +
    Rural_percent + Pct_65_older + PCT_18_younger + restaurants_per_capita +
    Household_burden + SNAP_low_access_pct *Recovered_Calories ,
  data = foodinsec_train,
  family = binomial
)
summary(model02)
```

```{r}



predictions <- predict(model1, newdata = foodinsec_test)

rmse_test <- rmse(foodinsec_test$Foodinsecurity, predictions)

rmse_test


```

```{r}
# model 2
set.seed(12345)
lm_model2 <- lm(
  Foodinsecurity ~ BlackPct + AsianPct + WomenWICpart + NativePct + HispPct + TwoOrMorePct +
                   Medianincome + low_income_low_store_access + adult_obese_rate + 
                   SummerFoodProgChange + WICChange + SNAP_redemptions + SNAP_per_capita +
                   WIC_redemptions + SNAP_auth_stores + SNAPstoreschange +
                    calories_per_person,
  data = foodinsec_train
)
summary(lm_model2)
```

```{r}
predictions2 <- predict(lm_model2, newdata = foodinsec_test)

rmse_test2 <- rmse(foodinsec_test$Foodinsecurity, predictions)

rmse_test2
```

```{r}
AIC(model1, lm_model2)
BIC(model1, lm_model2)
```

## Random Forest

```{r}
set.seed(111)
rf_model <- randomForest(
  Foodinsecurity ~ WhitePct + BlackPct + NativePct + 
    PacificIslanderPct +  Medianincome + NocarlowaccessChange + 
    ObesityPct + WICChange + SNAPparticipantschange + 
    SNAPstoreschang + SummerFoodProgChange + MetroStatus + 
    WomenWICpart,
  data = foodinsec_train,
  ntree = 500,
  importance = TRUE
)

print(rf_model)
importance(rf_model)
varImpPlot(rf_model)
```

```{r}
set.seed(1235)
control <- trainControl(method = "cv", number = 5)
rf_cv <- train(
  Foodinsecurity ~ WhitePct + BlackPct + NativePct + PacificIslanderPct +
    Medianincome + NocarlowaccessChange + ObesityPct + WICChange +
    SNAPparticipantschange + SNAPstoreschang + SummerFoodProgChange +
    MetroStatus + WomenWICpart,
  data = model_data,
  method = "rf",
  trControl = control,
  tuneLength = 3
)
print(rf_cv)
```

# Conclusion

The first model developed to predict food insecurity by county is a multiple linear regression which included key demographic variables, as well as economic, access to stores, and participation in food programs. This model scored an adjusted R\^2 of 71%. Most variables were statistically significant, except for the Asian population in the county and the adult obesity rate. As expected, median income, various types of food programs participation changes, and overall obesity rates were statistically significant. This indicates that some of the variables such as adult obesity rate was redundant and some multicollinearity.

The second model removed the weaker predictors, due to the insignificance to the model. However, the adjusted R\^2 remained about the same at \~71%. This concluded that the excluded variables were in fact not valuable to the model.

Lastly, the random forest model used the same predictors as the previous model to capture any nonlinear relationships among the variables. This modeled outperformed with an R\^2 of 99.97%, an almost perfect accuracy rate. WIC participation from women was the most influential variable, which is as expected because this variable had a high correlation with food insecurity.

```{r}
model_summary <- data.frame(
  Model = c("Linear Regression (Full)", 
            "Linear Regression (Reduced)", 
            "Random Forest"),
  RMSE = c(1.295, 1.29, 0.0385),
  R2 = c(0.7091, 0.7091, 0.99976),
  MAE = c(NA, NA, 0.0105)  
)

model_summary

```

```{r}
model_long <- model_summary %>%
  pivot_longer(cols = c(RMSE, R2), names_to = "Metric", values_to = "Value")


ggplot(model_long, aes(x = Model, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("R2" = "#1b9e77", "RMSE" = "#d95f02")) +
  labs(title = "Model Performance Comparison",
       x = "Model",
       y = "Metric Value") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
ggplot(model_long, aes(x = reorder(Model, -Value), y = Value, fill = Metric)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(title = "Model Comparison: Accuracy and Error",
       x = "Model", y = "Value") +
  theme_minimal(base_size = 13)
```
